---
title: "Reproducible research week 2 assignment"
output:
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the data

Set working directory and download and unzip the file from the internet.

```{r load}
# Set working directory
setwd("~/Documents/BGS_admin/Training/Data_science/Module_5_Reproducible Research/Week_2_not_completed/Assignment")

# Download the file from the internet
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", destfile = "activity.zip", mode="wb")

# Unzip the file
unzip("activity.zip")
Steps <- read.csv("activity.csv", header = TRUE)
```

## Summarise the data

Summarise the step data.

```{r summary}
# Summarise the data
head(Steps)
summary(Steps)
str(Steps)
```

## What is mean total number of steps taken per day?

The total number of steps taken per day is shown below. 
```{r steps}
# Aggregate the steps taken by day and then sum to calculate the total steps taken per day
Steps_Day <- aggregate(Steps$steps, by=list(Steps$date), sum)
colnames(Steps_Day) <- c("Day", "Total_steps")

# Display the result
head(Steps_Day)
```

Figure 1 shows a histogram of the total steps taken per day.
```{r steps histogram, echo=TRUE, fig.width=6, fig.height=4}
hist(Steps_Day$Total_steps, xlab = "Steps", breaks = 20, main="Total number of steps per day", col = c("blue"))
```

The mean and median of the total number of steps taken per day is shown below.
```{r summary stats}
# Calculate the mean
mean(Steps_Day$Total_steps, na.rm=TRUE)

# Calculate the median
median(Steps_Day$Total_steps, na.rm=TRUE)
```

## What is the average daily activity pattern?

Figure 2 shows a time series plot of the average number of steps across all days plotted against the whole day split into 5 minute intervals.
```{r steps timeseries, echo=TRUE, fig.width=6, fig.height=4}
# Create a new matrix with mean values for each 5 minute interval, in order to do so remove any 5 minute interval without data.
NA_ind <- is.na(Steps$steps)
Steps_nNA <- Steps[!NA_ind,]
Steps_int <- aggregate(Steps_nNA$steps, by=list(Steps_nNA$interval), mean)
colnames(Steps_int) <- c("Interval", "Steps")

# Plot the average number of steps across the days per 5 minute interval
plot(Steps_int$Interval, Steps_int$Steps, type = "l", xlab = "Five minute interval", ylab = "Average steps", main="Average steps per 5 minute interval")
```

The 5 minute interval during the day, which on average, has the maximum number of steps is shown below.
```{r max steps}
Steps_int_max <- Steps_int[which.max(Steps_int$Steps),]
Steps_int_max
```


## Imputing missing values

The number of missing values from the database are shown below.
```{r missing values}
No_vals <- sum(NA_ind)
No_vals
```

The missing values are replaced in the original dataset with the mean of each 5 minute interval calculated using the remainder of the dataset.
```{r replace missing values}
# Construct a function to replace the missing values
replace_with_mean <- function(num) replace(num, is.na(num), mean(num, na.rm = TRUE))

#Replace the missing values
library(dplyr)
Steps_replaced <- (Steps %>% group_by(interval) %>% mutate(steps = replace_with_mean(steps)))

# Display the new dataset
head(Steps_replaced)

# Check that no missing values remain.
sum(is.na(Steps_replaced))
```

The new dataset is summarised below.
```{r new dataset summary}
# Convert the new dataset to a dataframe
Steps_replaced <- as.data.frame(Steps_replaced)

# Summarise the new dataframe
head(Steps_replaced)
summary(Steps_replaced)
str(Steps_replaced)
```

The mean and median of the total number of steps taken per day is shown below.
```{r steps new dataset}
# Aggregate the steps taken by day and then sum to calculate the total steps taken per day
Steps_Day_rep <- aggregate(Steps_replaced$steps, by=list(Steps_replaced$date), sum)
colnames(Steps_Day_rep) <- c("Day", "Total_steps")

# Calculate the mean
mean(Steps_Day_rep$Total_steps)

# Calculate the median
median(Steps_Day_rep$Total_steps)

```
Replacing the missing values has increased the median by 1.19, but the mean remains the same.

Figure 3 shows a histogram of the total steps taken per day but using the new dataset (red) compared to the old dataset (blue).
```{r steps histogram new dataset, echo=TRUE, fig.width=6, fig.height=4}
par(1,2)
hist(Steps_Day_rep$Total_steps, xlab = "Steps", breaks = 20, main="Total number of steps per day", col = c("red"))
hist(Steps_Day$Total_steps, xlab = "Steps", breaks = 20, main="Total number of steps per day", col = c("blue"))
```

## Are there differences in activity patterns between weekdays and weekends?

The new dataset is split up into weekends and weekdays and activity levels are compared in Figure 4.
```{r steps timeseries new dataset, echo=TRUE, fig.width=6, fig.height=4}
# Create new dataframe wth weekday and weekend variables
Steps_replaced$date <- as.Date(Steps_replaced$date)
Steps_replaced$weekday <- weekdays(Steps_replaced$date)
Steps_replaced$weekend <- ifelse(Steps_replaced$weekday=="Saturday" | Steps_replaced$weekday=="Sunday", "Weekend", "Weekday" )

# Plot the difference between activity patterns on weekdays and weekends.
library(ggplot2)
Week_period <- aggregate(Steps_replaced$steps , by= list(Steps_replaced$weekend, Steps_replaced$interval), na.omit(mean))
names(Week_period) <- c("weekend", "interval", "steps")

ggplot(Week_period, aes(x=interval, y=steps, color=weekend)) + geom_line()+
    facet_grid(weekend ~.) + xlab("Interval") + ylab("Mean of Steps") +
    ggtitle("Comparison of Average Number of Steps in Each Interval") + 
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black"))
```

There is a difference in activity levels between weekdays and weekends, in particular there is a more gradual increase in activity at the weekend, and activity levels remain more consistent.